TacticToe: Learning to Prove with Tactics

### 2018-03-20 tactictoe_jv.tex
integrating Josef's comments wip

\def\holfour{\textsf{HOL4}\xspace}
\def\isabelle{\textsf{Isabelle}\xspace}
\def\hollight{\textsf{HOL Light}\xspace}
\def\mizar{\textsf{Mizar}\xspace}
\def\coq{\textsf{Coq}\xspace}
\def\ocaml{\textsf{OCaml}\xspace}
\def\vampire{\textsf{Vampire}\xspace}
\def\eprover{\textsf{E-prover}\xspace}
\def\zthree{\textsf{z3}\xspace}

@@

\begin{abstract}
We implement a tactical prover TacticToe on top of the HOL4 interactive theorem
prover. TacticToe learns from human proofs which mathematical technique is
the most suitable in each proof situation.
This knowledge is then used by a Monte Carlo tree search algorithm to
explore the most promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.4\% of
7164 theorems in HOL4's standard library whereas
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the
results of TacticToe and Eprover.
\end{abstract}

\tableofcontents

@@

\paragraph{Goals and theorems.}
A goal (or proof state) is represented as a \emph{sequent}.
A sequent is composed of a set of assumptions and a conclusion, all of which are higher-order logic formulas.
When a goal is proven, it becomes a theorem.
If the developer gives the theorem a name then we refer to it as a top-level theorem.

@@

\paragraph{Tactics in \holfour.}
Considering concrete implementation details now, a \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a validation function.
Calling the validation function on a list of \emph{theorems}, corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the validation
function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of backwards, goal-directed, proof
construction.
Because validation functions are only important to check the final proof, we
omit them during the description of the proof search algorithm. We denote by
$t(g)$ the list of goals produced by the application of a tactic $t$ to a goal
$g$.

\subsection{Contributions}
This paper extends the work described in \cite{tgckju-lpar17}.
There we proposed the idea of a tactic-based prover based on supervised
learning guidance. And we achieved a 39\% success rate on theorems of the
\holfour standard library by running \tactictoe with a 5 seconds timeout.
The additional contributions and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner.
\item A tactic abstraction mechanism is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\%
success rate on the standard library.
\item First evaluation of \tactictoe success rate relative to
the length of the original human proofs.
\item Minimization and embellishment of the generated proof  facilitates user
interaction.
\end{itemize}

@@

\end{tikzpicture}
\end{center}
\label{fig:components}
\caption{Relation between modules of the learning and proving tool chains.
Functions are represented by rectangles and objects by ellipses.}
\end{figure}

The different components depicted in Fig~\ref{fig:components} give a 
high-level overview of the techniques necessary to generate 
tactic-level proofs by learning from recorded knowledge. From a formal library 
of ITP proofs, we extract and record tactic invocations 
in Section~\ref{sec:recording}. Based on such training examples, we
  build efficient machine learning classifiers in Section~\ref{s:prediction}. 
  At the basis of those predictions are 
  similarity measures computed in Section~\ref{sec:features} from syntactic 
  features of mathematical formulas .
  Modification of the knowledge base performed by orthogonalization 
  in Section~\ref{sec:ortho} and abstraction in Section~\ref{sec:synthesis}
  improves the quality of the databases.
  And the preselection method described in Section~\ref{sec:presel} increases 
  the speed 
  of the prediction algorithm.
  The learned predictors relying on this knowledge serves as
  guidance for the proof search algorithm presented in 
  Section~\ref{sec:proofsearch} 
  which explores different proof paths in the proof tree defined in 
  Section~\ref{sec:prelim}. 
  The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics minimized in Section~\ref{sec:proofdisplay}. And in 
  Section~\ref{s:experiments}, we evaluate the system 
  on a large set of theorems originating from \holfour.
  And generated tactic-level proofs are analysed in 
  Section~\ref{sec:case_study}.

The section order does not follow the implementation. In this way, 
we can give a description of the proof search algorithm sooner. This is 
beneficial for the reader as it shows the essential for our work before 
technical details on the prediction and recording algorithms.

\section{Search Tree}\label{sec:prelim}
The application of a tactic to an initial conjecture produces a list of goals.
Each of the created goals can in turn be the input to other tactics.
Moreover, it is possible to try multiple tactics on the same goal.
In order to keep track of the progress made so far from the initial conjecture
we want to solve, we organize goals and tactics into a graph with list of goals
as nodes and tactics as edges (See Definition~\ref{def:stree}).

In this section, we only give a description of the graph at a given
moment of the search after a certain amount of tactic applications.
Exploration of the tree is directed by the MCTS algorithm in
Section~\ref{sec:proofsearch}, which is guided by prediction algorithms
presented in Section~\ref{sec:predictions}.

\begin{definition}\label{def:stree}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes representing lists of goals.
\item $T$ is a function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It
takes a goal and return the set of tactics already applied to this goal.
\item $G$ is a function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$.
It takes a node and returns the list of goals of this node.
\item $A$ is a function that takes a pair $(g,t)\in \mathbb{G} \times 
\mathbb{T}$ such as $t \in T(g)$ and 
returns a node $A(g,t)$ such as $t(g) = G(A(g,t))$. In other words, the output 
$t(g)$ is exactly the list of goals contained in $A(g,t)$.
\item It is acyclic, i.e. a node cannot be a strict descendant (see 
Definition~\ref{def:desc}) of itself.
\item There is exactly one node with no parents. This root node contains 
exactly one goal which is the initial goal (also called conjecture).
\end{itemize}

@@

If no explicit order is given, we assume that the sets
$\mathbb{T},\mathbb{G},\mathbb{A}$ are equipped with an
arbitrary order. In Figure~\ref{fig:choice}, a part of the a search tree is
depicted, with goals represented by circles, nodes by rectangles and tactics by
arrows.

@@

\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1b);
\draw[-to,thick] (0) to node[xshift=-4](t2){$t_j$} (2b);
\draw[-to,dotted,thick] (0) to (1m);
\draw[-to,dotted,thick] (0) to (2m);
\draw[-to,dotted,thick] (0) to node[xshift=15](t2){$t_m$} (3b);
\end{tikzpicture}
\end{center}
\caption{A node of a search tree $a_0$, the list of goals $g_0 \ldots g_n$ it
contains and the nodes $a_1 \ldots a_m$ derived from the application of the
tactics $t_1 \ldots t_m$ to $g_i$.}
\label{fig:choice}
\end{figure}

In the following, properties are defined in the context of a
search tree $\mathfrak{T}$. In Definition~\ref{def:solved}, we specify
different status for a goal: open, pending or solved. There, we define
what a solved goal and solved node are by mutual recursion on the number of
steps it takes to solve a goal and a node.
\begin{definition}\label{def:solved}(solved goal, solved nodes, open goal,
pending goal)\\
The set of solved nodes $\mathbb{A}^\infty$ and
the set of solved goals $\mathbb{G}^\infty$ are defined inductively by:

\begin{align*}
\mathbb{A}^{0} &=_{def}
\lbrace a \in \mathbb{A}\ |\ G(a) = \emptyset \rbrace \\
\mathbb{G}^{0} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists t \in T(g).\ A(g,t) \in \mathbb{A}^{0} \rbrace\\
\mathbb{A}^{n+1} &=_{def} \lbrace a \in \mathbb{A}\ |\
\forall g \in G(a).\ g \in \mathbb{G}^{n} \rbrace\\
\mathbb{G}^{n+1} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists t \in T(g).\ A(g,t) \in \mathbb{A}^{n+1} \rbrace \\
\mathbb{A}^\infty &=_{def} \bigcup_{n \in \mathbb{N}} \mathbb{A}^n \ \ \ \ \
\mathbb{G}^\infty =_{def} \bigcup_{n \in \mathbb{N}} \mathbb{G}^n\\
\end{align*}

We call a goal (respectively node) unsolved if it is not in the set of solved
goals (respectively nodes).
The open goal of a node is the first unsolved goal of this node according to
some preset order. The other unsolved nodes are called pending nodes.
If a node is unsolved, then it contains a unique open goal.
During MCTS exploration, we do not explore pending goals of an unsolved node
before its open goal is solved which justifies our terminology. In other words,
our search tree has the following property: if $g$ is a pending node then
$T(g)$ is empty.

@@

\begin{definition}\label{def:desc}(children, descendant, ancestors)\\
The children of a goal $g$ is a set of nodes defined by:
\[\mathit{Children}(g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace\]

By extension, we define the children of a node $a$ by:
\[\mathit{Children}(a) = \bigcup_{g \in G(a)} \mathit{Children}(g) \]

The descendants of $a$ by:
\begin{align*}
\mathit{Descendant}^{0}(a) &=_{def} \lbrace a \rbrace \\
\mathit{Descendant}^{n+1}(a) &=_{def} \bigcup_{a' \in \mathit{Descendant}^{n}(a)}
\mathit{Children}(a') \\
\mathit{Descendant}(a) &=_{def} \bigcup_{n \in \mathbb{N}} \mathit{Descendant}^n(a)\\
\end{align*}

And the ancestors of $a$ by:
\[\mathit{Ancestors}(a) =_{def} \lbrace b \in \mathbb{A} \ | \ a \in
Descendant(b) \rbrace\]

@@

\begin{definition}\label{def:prod} (productive tactic)
The application of a tactic on an open goal is called productive if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e., its output does not contain any goals that appear
in the ancestor of the node of the current goal.
\item It is not a parallel step, i.e., its output is not a superset
of a list of goals produced from the same goal.
\end{itemize}

@@

\section{Prediction}\label{s:prediction}
The learning-based selection of relevant lemmas significantly improves the
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe we use the
distance-weighted \emph{$k$ nearest-neighbour} classifier~\cite{DudaniS76}
adapted for lemma selection~\cite{ckju-pxtp13}. It allows for hundreds of
predictions per second, while maintaining very high prediction quality~\cite{femalecop}.
We rely on this supervised machine learning algorithm to predict three
different kinds of objects:
 tactics, theorems and lists of goals. The number of predicted objects $k$ is
 called the \emph{prediction radius}. For a goal $g$, we note
$\mathit{Predictions}^{\mathit{tactic}}_k (g)$ (respectively
$\mathit{Predictions}^{\mathit{theorem}}_k (g)$
and $\mathit{Predictions}^{\mathit{goal\_list}}_k (g)$) the $k$ tactics
(respectively theorems and lists of goals) selected by our prediction algorithm.

@@

\paragraph{Tactics}
To start with, we build a database of tactics consisting
of tactic-goal pairs recorded from successful tactic applications in human
proofs (recording will be discussed in
Section~\ref{sec:recording}).
During proof search, the recorded goals will be reordered according to their
similarity with a target goal $g$ (usually the first pending goal of a node).
The recorded pairs induce an
order on tactics. The intuition is that tactics which have been successful on
goals similar to $g$ should be tried first, since they are more likely to lead
to a proof.
This predicted tactic order is translated into a prior policy for our MCTS
algorithm (see Section~\ref{sec:policy}).
Tactic selection is also used to improve the quality of the database
of tactics during orthogonalization (see Section~\ref{sec:ortho}).

@@

\paragraph{Lists of Goals as Output of Tactics}
A dataset of tactic outputs is compiled during orthogonalization.
This set is separated into positive and negative examples (see
Section~\ref{sec:evaluation}).
List of goals that are a subset of the result of a recorded tactic on a
target goal are considered positive. During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

@@

Despite the best efforts of the prediction algorithms, a selected tactic may
not solve the current goal, proceed in the wrong direction
or even loop. For that reason, predictions need to be accompanied by a
proof search mechanism that allows for backtracking and
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search~\cite{montecarlo} algorithm
that relies
on a prior evaluation function and a prior policy function. Those priors are
learned through direct supervised learning for the policy and via the data
accumulated during orthogonalization for the evaluation. Therefore, this MCTS
algorithm does not need to rely on roll outs for evaluation. The
policy and evaluation are estimated by the simple supervised k-NN algorithm
here instead of neural networks trained through reinforcement learning in
AlphaGo Zero~\cite{silver2017mastering}.
A characteristic of the MCTS algorithm is that it offers a good trade-off
between exploitation and exploration. This means that the algorithm searches
deeper more promising branches and leaves enough time for the exploration
of less likely alternatives.

@@

If the highest children value is higher than the widening policy, then the
selected child
becomes the current node, otherwise the final result of node selection is the
current node. The node selection algorithm can be written as the following
pseudo-code:

@@

The number $\mathit{Failure}(a_i)$ is the amount of failures that occurred
during node extension from descendants of $a_i$ .
The current evaluation $CurEvaluation(a_i)$ is the average evaluation of
all descendants of $a_i$ including node extension failures.

@@

\[\mathit{CurPolicy}(a_i) = \frac{1 +
\mathit{Visit}(a_i))}{\sqrt{\mathit{Visit}(p))}}\]
The policy is approximately the  percentage of time a node was visited. The
square root skews this probability to favor even more exploration of nodes with
few visits.

@@

\section{Feature Extraction and Similarity Measure}\label{sec:features}
%We explain the construction of the similarity measure by which relevant
%objects are predicted

@@

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{lc}
\toprule
  & Solved (60s) \\
\midrule
   \eprover   & 2472 (34.5\%)\\
   \tactictoe & 4760 (66.4\%)\\
\midrule
   Total  & 4946 (69.0\%)\\
\bottomrule
\end{tabular}
\caption{Evaluation on 7164 top-level theorems of the \holfour standard library
\label{tab:_param}}
\end{table}

@@

\begin{table}[]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} & {meas} \\
\midrule
\tactictoe & 81.2 & 74.0 & 79.6 & 31.3\\
\eprover & 59.9 & 72.0 & 67.1 & 12.8\\
\midrule
\phantom{abc} & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 45.8 & 79.5 & 65.3 & 82.0 \\
\eprover & 24.1 & 26.5 & 15.8 & 24.7 \\
\bottomrule
\end{tabular}
\caption{\label{theories}Percentage (\%) of re-proved theorems in the theories
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}.
}
\end{table}

>>>>

\def\holfour{\textsf{HOL4}\xspace}
\def\isabelle{\textsf{Isabelle}\xspace}
\def\hollight{\textsf{HOL Light}\xspace}
\def\mizar{\textsf{Mizar}\xspace}
\def\coq{\textsf{Coq}\xspace}
\def\ocaml{\textsf{OCaml}\xspace}
\def\vampire{\textsf{Vampire}\xspace}
\def\eprover{\textsf{Eprover}\xspace}
\def\zthree{\textsf{z3}\xspace}

@@

\begin{abstract}
We implement a tactical prover TacticToe on top of the HOL4 interactive theorem
prover. TacticToe learns from human proofs which mathematical technique is
the most suitable in each proof situation.
This knowledge is then used by a Monte Carlo tree search algorithm to
explore the most promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.4\% of
7164 theorems in HOL4's standard library whereas
Eprover with auto-schedule solves 34.5\%. The success rate rises to 69.0\% by 
combining the results of TacticToe and Eprover.
\end{abstract}

@@

\paragraph{Goals and theorems.}
A goal (or proof state) is represented as a \emph{sequent}.
A sequent is composed of a set of assumptions and a conclusion, all of which 
are higher-order logic formulas~\cite{gordon01hol}.
When a goal is proven, it becomes a theorem.
If the developer gives the theorem a name then we refer to it as a 
\textit{top-level} theorem.

@@

\paragraph{Tactics in \holfour.}
Considering concrete implementation details now, a \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a \textit{validation function}.
Calling the validation function on a list of \emph{theorems}, 
corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the 
validation
function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of backwards, goal-directed, proof
construction.
Because validation functions are only important to check the final 
proof, we
omit them during the description of the proof search algorithm. We denote by
$t(g)$ the list of goals produced by the application of a tactic $t$ to a goal
$g$.

\subsection{Contributions}
This paper extends the work described in \cite{tgckju-lpar17}.
There we proposed the idea of a tactic-based prover based on supervised
learning guidance. And we achieved a 39\% success rate on theorems of the
\holfour standard library by running \tactictoe with a 5 seconds timeout.
The contributions of this extension and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise in 
Section~\ref{sec:recording}. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm in Section~\ref{sec:proofsearch}. The MCTS algorithm gives a 
more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors presented in 
Section~\ref{s:prediction} for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner in Section~\ref{sec:ortho}.
\item A tactic abstraction mechanism is introduced in 
Section~\ref{sec:synthesis}. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover in Section~\ref{sec:atp} to help \tactictoe 
during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\%
success rate on the standard library. And detailed comparisons between 
\tactictoe and \eprover on the different type of problems are reported in 
Section~\ref{s:experiments}.
\item Minimization and embellishment of the discovered proof facilitates user
interaction in Section~\ref{sec:proofdisplay}.
\end{itemize}

@@

\end{tikzpicture}
\end{center}
\caption{\label{fig:components}Relation between modules of the learning and 
proving tool chains. Functions are represented by rectangles and objects by 
ellipses.}
\end{figure}

The different components depicted in Fig~\ref{fig:components} give a 
high-level overview of the techniques necessary to generate 
tactic-level proofs by learning from recorded knowledge. From a formal library 
of ITP proofs, we extract and record tactic invocations 
in Section~\ref{sec:recording}. Based on such training examples, we
  build efficient machine learning classifiers in Section~\ref{s:prediction}. 
  At the basis of those predictions are 
  similarity measures computed from syntactic features of mathematical formulas 
  .
  Modification of the knowledge base performed by orthogonalization 
  in Section~\ref{sec:ortho} and abstraction in Section~\ref{sec:synthesis}
  improves the quality of the databases.
  And the preselection method described in Section~\ref{sec:presel} increases 
  the speed 
  of the prediction algorithm.
  The learned predictors relying on this knowledge serves as
  guidance for the proof search algorithm presented in 
  Section~\ref{sec:proofsearch} 
  which explores different proof paths in the proof tree defined in 
  Section~\ref{sec:prelim}. 
  The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics minimized in Section~\ref{sec:proofdisplay}. And in 
  Section~\ref{s:experiments}, we evaluate the system 
  on a large set of theorems originating from \holfour.
  The generated proofs are compared with human proofs in 
  Section~\ref{sec:case_study}.

The section order does not follow the implementation. In this way, 
we can give a description of the proof search algorithm sooner. This order is 
beneficial for the reader as it shows the essential for our work before 
technical details on the prediction and recording algorithms.

\section{Search Tree}\label{sec:prelim}
The application of a tactic to an initial conjecture produces a list of goals.
Each of the created goals can in turn be the input to other tactics.
Moreover, it is possible to try multiple tactics on the same goal.
In order to keep track of the progress made so far from the initial conjecture
we want to solve, we organize goals and tactics into a graph with lists of goals
as nodes and tactics as edges (See Definition~\ref{def:stree}).

In this section, we only give a description of the graph at a given
moment of the search after a certain amount of tactic applications.
Construction of the tree is done by the MCTS algorithm in
Section~\ref{sec:proofsearch}, which is guided by prediction algorithms
presented in Section~\ref{s:prediction}.

\begin{definition}\label{def:stree}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes representing lists of goals. All objects 
 are tagged with their position in the tree (see Figure~\ref{fig:choice}).
\item $T$ is a function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It
takes a goal and return the set of tactics already applied to this goal.
\item $G$ is a function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$.
It takes a node and returns the list of goals of this node.
\item $A$ is a function that takes a pair $(g,t)\in \mathbb{G} \times 
\mathbb{T}$ such as $t \in T(g)$ and 
returns a node $A(g,t)$ such as $t(g) = G(A(g,t))$. In other words, the output 
$t(g)$ is exactly the list of goals contained in $A(g,t)$.
\item It is acyclic, i.e. a node cannot be a strict descendant (see 
Definition~\ref{def:desc}) of itself.
\item There is exactly one node with no parents. This root node contains 
exactly one goal which is the initial goal (also called conjecture).
\end{itemize}

@@

If no explicit order is given, we assume that the sets
$\mathbb{T},\mathbb{G},\mathbb{A}$ are equipped with an
arbitrary \textit{total order}. In Figure~\ref{fig:choice}, a part of the a 
search tree is depicted, with goals represented by circles, nodes by rectangles 
and tactics by arrows.

@@

\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1b);
\draw[-to,thick] (0) to node[xshift=-4](t2){$t_j$} (2b);
\draw[-to,dotted,thick] (0) to (1m);
\draw[-to,dotted,thick] (0) to (2m);
\draw[-to,dotted,thick] (0) to node[xshift=15](t2){$t_m$} (3b);
\end{tikzpicture}
\end{center}
\caption{\label{fig:choice}A node of a search tree $a_0$, the list of goals 
$g_0 \ldots g_n$ it
contains and the nodes $a_1 \ldots a_m$ derived from the application of the
tactics $t_1 \ldots t_m$ to $g_i$.}
\end{figure}

In the following, properties are defined in the context of a
search tree $\mathfrak{T}$. In Definition~\ref{def:solved}, we specify
different status for a goal: open, pending or solved. There, we define
what a solved goal and solved node are by mutual recursion on the number of
steps it takes to solve a goal and a node.
\begin{definition}\label{def:solved}(solved goal, solved nodes, open goal,
pending goal)\\
The set of \textit{solved nodes} $\mathbb{A}^*$ and
the set of \textit{solved goals} $\mathbb{G}^*$ are defined inductively by:

\begin{align*}
\mathbb{A}^{0} &=_{def}
\lbrace a \in \mathbb{A}\ |\ G(a) = \emptyset \rbrace \\
\mathbb{G}^{0} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists t \in T(g).\ A(g,t) \in \mathbb{A}^{0} \rbrace\\
\mathbb{A}^{n+1} &=_{def} \lbrace a \in \mathbb{A}\ |\
\forall g \in G(a).\ g \in \mathbb{G}^{n} \rbrace\\
\mathbb{G}^{n+1} &=_{def} \lbrace g \in \mathbb{G}\ |\
\exists t \in T(g).\ A(g,t) \in \mathbb{A}^{n+1} \rbrace \\
\mathbb{A}^* &=_{def} \bigcup_{n \in \mathbb{N}} \mathbb{A}^n \ \ \ \ \
\mathbb{G}^* =_{def} \bigcup_{n \in \mathbb{N}} \mathbb{G}^n\\
\end{align*}

We call a goal (respectively node) \textit{unsolved} if it is not in the set of 
solved
goals (respectively nodes).
The \textit{open goal} of a node is the first unsolved goal of this node 
according to
some preset order. The other unsolved nodes are called pending nodes.
If a node is unsolved, then it contains a unique open goal.
During MCTS exploration, we do not explore pending goals of an unsolved node
before its open goal is solved which justifies our terminology. In other words,
our search tree has the following property: if $g$ is a pending node then
$T(g)$ is empty.

@@

\begin{definition}\label{def:desc}(children, descendant, ancestors)\\
The \textit{children of a goal} $g$ is a set of nodes defined by:
\[\mathit{Children}(g) = \lbrace A(g,t)\in \mathbb{A}\ |\ t \in T(g) \rbrace\]

By extension, we define the \textit{children of a node} $a$ by:
\[\mathit{Children}(a) = \bigcup_{g \in G(a)} \mathit{Children}(g) \]

The \textit{descendants of a node} $a$ by:
\begin{align*}
\mathit{Descendant}^{0}(a) &=_{def} \lbrace a \rbrace \\
\mathit{Descendant}^{n+1}(a) &=_{def} \bigcup_{a' \in \mathit{Descendant}^{n}(a)}
\mathit{Children}(a') \\
\mathit{Descendant}(a) &=_{def} \bigcup_{n \in \mathbb{N}} \mathit{Descendant}^n(a)\\
\end{align*}

And the \textit{ancestors of a node} $a$ by:
\[\mathit{Ancestors}(a) =_{def} \lbrace b \in \mathbb{A} \ | \ a \in
Descendant(b) \rbrace\]
\end{definition}

@@

\begin{definition}\label{def:prod} (productive tactic)
The application of a tactic $t$ on an open goal $g$ is called 
\textit{productive} if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e., its output does not contain any goals that appear
in the ancestor of the node of the current goal.
\item It is not a parallel step, i.e., there does not exist $t'$ in $T(g)$ 
such as $t'(g) \subseteq t(g)$.
\end{itemize}

@@

\section{Prediction}\label{s:prediction}
The learning-based selection of relevant lemmas significantly improves the
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe we use the
distance-weighted \emph{$k$ nearest-neighbour} classifier~\cite{DudaniS76}
adapted for lemma selection~\cite{ckju-pxtp13}. It allows for hundreds of
predictions per second, while maintaining very high prediction quality~\cite{femalecop}.
We rely on this supervised machine learning algorithm to predict three
different kinds of objects:
 tactics, theorems and lists of goals. The number of predicted objects $k$ is
 called the \emph{prediction radius}. For a goal $g$, we denote
by $\mathit{Predictions}^{\mathit{tactic}}_k (g)$ (respectively
$\mathit{Predictions}^{\mathit{theorem}}_k (g)$
and $\mathit{Predictions}^{\mathit{goal\_list}}_k (g)$) the $k$ tactics
(respectively theorems and lists of goals) selected by our prediction algorithm.

@@

\paragraph{Tactics}
To start with, we build a database of tactics consisting
of tactic-goal pairs recorded from successful tactic applications in human
proofs (recording will be discussed in
Section~\ref{sec:recording}).
During proof search, the recorded goals will be reordered according to their
similarity with a target goal $g$ (usually the first pending goal of a node).
The recorded pairs and their similarity to $g$ induce an
order on tactics. The intuition is that tactics which have been successful on
goals similar to $g$ should be tried first, since they are more likely to lead
to a proof.
This predicted tactic order is translated into a prior policy for our MCTS
algorithm (see Section~\ref{sec:policy}).
Tactic selection is also used to improve the quality of the database
of tactics during orthogonalization (see Section~\ref{sec:ortho}).

@@

\paragraph{Lists of Goals as Output of Tactics}
A dataset of tactic outputs is compiled during orthogonalization (see 
Section~\ref{sec:ortho}).
Some elements of this set will be considered positive examples (see 
Section~\ref{sec:evaluation}).
Lists of goals that are a subset or equal to the result of a recorded tactic on 
a target goal are considered positive because if a set of goals is solvable 
then a subset of it is too. During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

\subsection{Feature Extraction and Similarity Measures}\label{sec:features}

@@

Despite the best efforts of the prediction algorithms, a selected tactic may
not solve the current goal, proceed in the wrong direction
or even loop. For that reason, predictions need to be accompanied by a
proof search mechanism that allows for backtracking and
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search~\cite{montecarlo} algorithm
that relies
on a prior evaluation function and a prior policy function. Those priors are
learned through direct supervised learning for the policy and via the data
accumulated during orthogonalization for the evaluation. Therefore, this MCTS
algorithm does not need to rely on roll outs for evaluation. The
policy and evaluation are estimated by the simple supervised k-NN algorithm
here instead of neural networks trained through reinforcement learning
as e.g. in AlphaGo Zero~\cite{silver2017mastering}.
A characteristic of the MCTS algorithm is that it offers a good trade-off
between exploitation and exploration. This means that the algorithm searches
deeper the more promising branches and leaves enough time for the exploration
of less likely alternatives.

@@

If the highest children value is higher than the \textit{widening policy} (see 
Section~\ref{sec:policy}), then the
selected child
becomes the current node, otherwise the final result of node selection is the
current node. The node selection algorithm can be written as the following
pseudo-code:

@@

The number $\mathit{Failure}(a_i)$ is the number of failures that occurred
during node extension from descendants of $a_i$ .
The current evaluation $CurEvaluation(a_i)$ is the average evaluation of
all descendants of $a_i$ including node extension failures.

@@

\[\mathit{CurPolicy}(a_i) = \frac{1 +
\mathit{Visit}(a_i)}{\sqrt{\mathit{Visit}(p)}}\]
The policy is approximately the  percentage of time a node was visited. The
square root skews this probability to favor even more exploration of nodes with
few visits.

@@

@@

\begin{table}[h!]
\centering\ra{1.3}
\small
\begin{tabular}{lc}
\toprule
  & Solved (60s) \\
\midrule
   \eprover   & 2472 (34.5\%)\\
   \tactictoe & 4760 (66.4\%)\\
\midrule
   Total  & 4946 (69.0\%)\\
\bottomrule
\end{tabular}
\caption{\label{tab:_param} Evaluation on 7164 top-level theorems of the 
\holfour standard library
}
\end{table}

@@

\begin{table}[]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} & {meas} \\
\midrule
\tactictoe & 81.2 & 74.0 & 79.6 & 31.3\\
\eprover & 59.9 & 72.0 & 67.1 & 12.8\\
\midrule
\phantom{abc} & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 45.8 & 79.5 & 65.3 & 82.0 \\
\eprover & 24.1 & 26.5 & 15.8 & 24.7 \\
\bottomrule
\end{tabular}
\caption{\label{theories} Percentage (\%) of re-proved theorems in the theories
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}.
}
\end{table}
